---
title: "Data QA/QC Notebook on UNCORRECTED Raw Data"
output: html_notebook
---
# Introduction
This notebook carries out some basic data QA/QC checks on the raw data. 
This version runs the QA/QC checks on the uncorrected data, to demonstrate the
process used to identify problems.  A second version of this notebook runs on
the corrected data, to demonstrate the effect of the data correction process.

Note that this is based on analyzing the data already reorganized according to
the code in the `Data Reorganization UNCORRECTED.Rmd` R notebook.

This version is outdated, and no longer being fully maintained but retained to 
help document data correction.

# Import Libraries
```{r}
library(readr)
library(tidyverse)
```

# Import Data
Note that we are loading an **uncorrected** file, not drawing directly from the
Excel Source File.  This file has been lightly processed to extract data from
the Excel File, via the `Data_Reorganization_UNCORRECTED.Rmd` notebook.
```{r}
fn <- 'presumpscot_UNCORRECTED.csv'

presumpscot_data <- read_csv(fn,
    col_types = cols(Ecoli = col_double(),
                     X1 = col_skip()))

presumpscot_data <- presumpscot_data %>%
  select (-Time) %>%
  mutate(Month = factor(as.numeric(format(Date, '%m')),
                        labels = month.abb[5:9]))  # Note that as.numeric()
                                                   # will not give month number
```

# Data Prevalence
## *E.coli* data by year
```{r}
presumpscot_data %>%
  group_by(Site, Year)%>%
  summarize(n=sum(! is.na(Ecoli)))%>%
  spread(Year, n)
```

## DO data by year
```{r}
presumpscot_data %>%
  group_by(Site, Year)%>%
  summarize(n=sum(! is.na(DO)))%>%
  spread(Year, n)
```
Apparently, a total of 47 locations have been sampled by PRW since 2009. The
normal sampling plan in each year is to sample a fixed number of Saturdays,
usually eight.  We often see more than eight sample records at a site within a
year.  Generally this is because QA/QC replicate samples were collected, or DO
data were collected at different depths on one sample date.  Multiple depth
samples were only collected in 2011, 2012, and 2013.

Prior analysis has shown little effect of sample depth.  There sometimes are
substantial differences in observed E. coli values for replicate samples,
expressed as a percent difference.  On a log scale, however, the differences are
less apparent.

## Create a "surface water only"" data set
Note that this is not necessary for E.Coli, but is for DO.
```{r}
surface_data <- presumpscot_data %>%
  filter(is.na(Depth) | Depth == 0)
```

### Review number of Samples
```{r}
cat('All Data\n')
a <- with(presumpscot_data, cbind(length(Ecoli),
                             sum(! is.na(Ecoli)),
                             sum(! is.na(DO))))
colnames(a) <- c('N', 'E.coli', 'DO')
a

cat('Surface Data\n')
a <- with(surface_data, cbind(length(Ecoli),
                         sum(! is.na(Ecoli)),
                         sum(! is.na(DO))))
colnames(a) <- c('N', 'E.coli', 'DO')
a
rm(a)
```
We lose approximately 100 samples, all affecting DO samples, not *E. coli*
samples.

# Analysis of duplicates
This is for QA/QC purposes.  We may want to look at variation among duplicates.
```{r}
duplicates = presumpscot_data %>%
  group_by(Date, Site) %>%
  mutate(nColi= sum(!is.na(Ecoli))) %>%
  ungroup() %>%
  filter(nColi>1) %>%
  mutate(SampleID = paste0(Site,'.',Date)) %>%
  arrange(Date) %>%
  select(SampleID, Site, Date, Year, Month, Ecoli, Flag, DO, PctSat)
```


```{r}
diffs <- duplicates %>%
  group_by(SampleID) %>%
  summarize(Year = min(Year),
            m = (max(Ecoli, na.rm=TRUE)+min(Ecoli, na.rm=TRUE))/2,
            d= max(Ecoli, na.rm=TRUE)-min(Ecoli, na.rm=TRUE),
            pctd = d/m)
  summary(diffs)
```
So, that shows that there are sometimes differences that are extremely large.  A
better analysis might be of logs of abundance, given the heavy skew of the
observed data.
```{r}
ns = diffs %>%
  group_by(Year) %>%
  summarize(n=n()) %>%
  mutate(n2 = paste("N=", n))


ggplot(diffs, aes(factor(Year), pctd)) +
  geom_boxplot() + 
  geom_text(aes(label=n2, x=factor(Year), y=1.75 ), data=ns)
```
So, the most extreme variation came from year 2017, and there were no duplicates 
in 2018, and few in 2019.

Given the skewed nature of these data, these high percent differences are not
dreadful, but they don't make anyone happy either.  A percent difference of more
than 150%, as observed three times in 2017, is problematic.

But notice the differences observed in 2019:
```{r}
diffs %>%
  filter(Year==2019) %>% pull(d)
```
That is clearly an error of some kind.  I eventually corrected those, which
appear to be spurious duplicate E. coli observations, by removing one of each
pair of identical E. coli observations, while leaving in place the different DO
observations reported.

# Data Evaluation
## E Coli QA/QC Review
### Histograms
```{r}
plt <- ggplot(presumpscot_data, aes(Ecoli)) +
  geom_histogram(aes(fill = Flag), bins=50) #+
  #geom_density()
plt
```
Notice that the values, while conceptually continuous are converted to select
discrete values by the MPN method used. Technically this is interval censored
data, but there are so many categories that the impact on analysis is probably
small.

The number of "uncensored" large values looks implausibly high, and a close look
shows that several slightly different values are all included with NEARLY
identical values.  I suspect that reflects different coding practices over time.
Again, differences are too small to matter for any quantitative analysis, so we
leave all observations as is.

```{r}
plt  + scale_x_log10()
```
We have a fat tailed mound-like distribution, which suggests a lognormal
distribution may work well for these data.  But note also that we have an excess
of elevated values, if we think this follows a lognormal distribution.  An
alternative is a Pareto Distribution, which also fits these data fairly well.
The Pareto predicts more extreme values than the lognormal. (see the analysis
notebooks for more details).

# Plotting Relationship with DO
We are looking here for any outliers.
```{r}
plt <- ggplot(surface_data, aes(DO, Ecoli)) +
  geom_point(aes(color=factor(Year))) +
  geom_smooth() +
  scale_y_log10()
plt
```
Other than a few extreme DO values, there is nothing obviously amiss here.
There is a weak indication that higher DO is associated with lower E. coli, as
might be expected if summer low flows and high temperatures are bad for both
dissolved oxygen and bacteria loads. We can do little more with this based on
these data because we lack temperature data for 2018 and 2019.

## DO QA/QC Review
```{r}
plt <- ggplot(surface_data, aes(DO)) + geom_histogram()
plt
```
We notice a few implausible low and a few implausible high DO observations.

```{r}
plt <- ggplot(surface_data, aes(DO, PctSat)) +
  geom_point(aes(color =factor(Year))) +
  geom_smooth()
plt
```
Note a handful of observations that are far from the trendline. Most are from
2018.  A couple are also from 2014, and one from 2019. Lets explore those.

```{r}
anomolies <- surface_data %>%
  filter((PctSat>75 & (DO<6 | DO>13)) | (PctSat<25 & DO>4 ) | PctSat>115) %>%
  select(-Name, -QC, -Depth, -Flag, -Month)
anomolies
#write.csv(anomolies,"anomolies.csv")

```
Careful review shows many of these represent implausible of DO and Percent
Saturation.  I sent a modified list of likely errors to Toby, who passed them on
to DEP, who returned suggested corrections to the 2018 data, based on DEP's
QA/QC procedures.